Computing likelihoods under the SMC for a general class of ARGs.

### 2024-04-22 paper.tex
update figures

\comment{Adapt figure to represent this}

@@

So, an ARG can be thought of either as a collection of inheritance relationships
between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $a$
inherited the left half of its genome from ancestor 6
and the right half from ancestor 5
is seen in the fact that the first two trees have $a$ below 6,
and the remaining have $a$ below 5.
The blue dotted line from $a$ in the second tree to 5 in the third tree
draws our attention to the fact that
the path traced by $a$'s ancestry switches at that point.
In fact, $a$ is in blue because it is a \emph{recombination} node --
\comment{wait is it ``recombination node'' or ``recombinant node''?}
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labeled by black numbers) are either samples
or \emph{coalescence} nodes --
ancestors that are the most recent common ancestor for two samples
at some place along the genome.

Crucial for the equivalence between both representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}.
Recombination nodes (in blue) have a single child in all trees.
However, coalescent nodes can be locally unary as well:
for instance, sample 3 inherited a longer stretch of genome from ancestor 5
than did sample 2.
In the top plot, this is visible because the pink bar below node 5
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node 5 is parent to 3
in the second tree, but not to node 2, and so is unary.

A key observation of \citet{mcvean_approximating_2005}
was that the sequence of trees along the genome
generated by the SMC was Markovian \emph{after removing all unary nodes}.
In other words, if we remove recombination nodes entirely,
and erase the unary portions of coalescent nodes,
then the distribution of the next tree can be generated only knowing the current tree.
(To see that the unary nodes makes the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree, we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
However, removing these unary portions of nodes seems odd
when one considers the consequences in the haplotype inheritance graph
of the top figure.
As pictured, both figures tell us that
3 inherits the middle bit of the genome from 7 through 5.
However, if we erase 5 from the second tree,
does this suggest that 3 inherited a single segment from 7 along two distinct paths?
The haplotype inheritance graph above makes it clear that this is unlikely:
it is in fact reasonable to leave in these unary sections.

Furthermore, these unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombination nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombination nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombination node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014}
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

Finally, acknowledging the uncertainty associated with ARG inference does not
necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
the ARG for which we lack information for confident reconstruction and resolving
them stochastically by sampling from the coalescent rather than completely arbitrarily
would already be a major step forward. 
% this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
Furthermore, we believe that quantifying uncertainty
would not necessarily always have to be associated with sampling from the space of all
compatible genealogies. Uncertainty could also
be quantified and passed on as metadata to the nodes in the data format storing the ARG.

@@

Next, we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
The tree sequence format encodes relationships using \emph{edges}
just as described above,
along with additional structure that
allows us to sequentially recover the local trees
along the genome very efficiently, and to reason about
the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the required strict total ordering on edges
(see definition \ref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.

>>>>

@@

So, an ARG can be thought of either as a collection of inheritance relationships
between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $a$
inherited the left half of its genome from ancestor 6
and the right half from ancestor 5
is seen in the fact that the first two trees have $a$ below 6,
and the remaining have $a$ below 5.
The blue dotted line from $a$ in the second tree to 5 in the third tree
draws our attention to the fact that
the path traced by $a$'s ancestry switches at that point.
In fact, $a$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labeled by black numbers) are either samples
or \emph{coalescence} nodes --
ancestors that are the most recent common ancestor for two samples
at some place along the genome.

Crucial for the equivalence between both representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}.
Recombinant nodes (in blue) have a single child in all trees.
However, coalescent nodes can be locally unary as well:
for instance, sample 3 inherited a longer stretch of genome from ancestor 5
than did sample 2.
In the top plot, this is visible because the pink bar below node 5
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node 5 is parent to 3
in the second tree, but not to node 2, and so is unary.

A key observation of \citet{mcvean_approximating_2005}
was that the sequence of trees along the genome
generated by the SMC was Markovian \emph{after removing all unary nodes}.
In other words, if we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes,
then the distribution of the next tree can be generated only knowing the current tree.
(To see that the unary nodes makes the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree, we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
However, removing these unary portions of nodes seems odd
when one considers the consequences in the haplotype inheritance graph
of the top figure.
As pictured, both figures tell us that
3 inherits the middle bit of the genome from 7 through 5.
However, if we erase 5 from the second tree,
does this suggest that 3 inherited a single segment from 7 along two distinct paths?
The haplotype inheritance graph above makes it clear that this is unlikely:
it is in fact reasonable to leave in these unary sections.

Furthermore, these unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014}
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

Finally, acknowledging the uncertainty associated with ARG inference does not
necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
the ARG for which we lack information for confident reconstruction and resolving
them stochastically by sampling from the coalescent rather than completely arbitrarily
would already be a major step forward.

% this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
We believe that quantifying uncertainty does not have to be associated with sampling
from the space of all compatible genealogies. Uncertainty could also
be quantified and passed on as metadata to the nodes in the data format storing the ARG.

@@

Here, we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
The tree sequence format encodes relationships using \emph{edges}
just as described above,
along with additional structure that
allows us to sequentially recover the local trees
along the genome very efficiently, and to reason about
the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the required strict total ordering on edges
(see definition \ref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.
